\bf Comparing the Power of Plot Designs to Reveal Correlation

### 2016-01-20 paper/paper.Rnw
nearly done with intro

The law then has two effects. If positive correlation exists in temporal data, the observer may be drawn to it faster from an overlaid line graph display. If positive correlation does not exist, or only exists ephemerally, the observer may be fooled by the overlaid line graphs into thinking there is strong association. This latter effect is a problem.

@@

There have been a few studies related to plotting temporal data.  \citet{Robbins} writes a blog and consults on plotting financial data. She is critical of the overlaid line graph, because it is misleading if the purpose is to read the difference between the two series. Her arguments date back a long time to \citet{playfair,tufte,friendly}. It is well-established that we cannot read differences between two series accurately, that the difference should be plotted separately if that is the information of interest.

\citet{Vanderplas} examines a related problem, dubbed the `sine illusion', where the human eye is poor at interpreting lines with sharp slopes. For time series, this means that people do not effectively read variance of seasonal series.

Correlation study by the ... guys

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{murders}
  \caption{\citep{Robbins} Murders in New York State (Orange) and New York City (Blue). The difference between the two represents murders outside of the city. What size would you judge the difference to be in 1994?}
  \label{murder}
\end{figure}

@@

>>>>

The law then, has two effects. If positive correlation exists in temporal data, the observer may be drawn to it faster from an overlaid line graph display. If positive correlation does not exist, or only exists ephemerally, the observer may be fooled by the overlaid line graphs into thinking there is strong association. This latter effect is a problem.

@@

There have been a few studies related to plotting temporal data.  \citet{Robbins} writes a blog and consults on plotting financial data. She is critical of the overlaid line graph, because it is misleading if the purpose is to read the difference between the two series. Her arguments date back a long time to \citet{playfair,tufte,friendly}. It is well-established that we cannot read differences between two series accurately, that the difference should be plotted separately if that is the information of interest. \citet{Vanderplas} examines a related problem, dubbed the ``sine illusion'', where the human eye is is fooled into erroroneously thinking that variation is larger at the peaks and troughs of seasonal time series. \citet{Vanderplas} developed a mathematical correction to the illusion.

Reading correlation between two sets of non-temporal context numbers has been examined in many papers. In \citet{harrison2014ranking} different graphical displays of two sets of numbers are ranked of their effectiveness by Weber's Law \citep{webers} which provides a relationship between perception and the ``effect size'' in the data. For example, two series with correlation 0.9 should be proportionately easier to distinguish from noise than two series with correlation 0.3. The visual display should reflect this difference is ease. Early papers \citep{Cleveland1982,lane1985} showed that people read association from a scatterplot differently than they would linearly interpret a correlation coefficient.

@@

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=0.5\textwidth]{murders}
%  \caption{\citep{Robbins} Murders in New York State (Orange) and New York City (Blue). The difference between the two represents murders outside of the city. What size would you judge the difference to be in 1994?}
%  \label{murder}
%\end{figure}

@@

Visual Inference background next
